Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2019-12-23 analysis/paper/paper.Rmd
work on PCA results

`Post-European` <- mshp$poe %T>% coo_plot( border="blue")
`Pre-European` <-  mshp$pre %T>% coo_plot(border="red", plot.new=FALSE)
`Chinese` <-       mshp$chi %T>% coo_plot(border="green", plot.new=FALSE)
legend("topright", 
       lwd=1,
       col=c("red", "blue", "green"), 
       legend=c( "Pre-European", "Post-European", "Chinese"))
```

@@

# Doesn't work if the plot is not produced by ggplot?
# plot_two_de_by_period <- 
  #plot_grid(, labels = c('A', 'B'))

@@

```{r PCA-table}
# these are fast
op.p <- PCA(op)
# proportion of variances explained, PC1 and PC2 explain 79% of variation
pro_vari <-scree_plot(op.p)$data
vari_two <- round(pro_vari$cumsum[2],4)*100
vari_three <- round(pro_vari$cumsum[3],4)*100
```

Outlines generated by EFA are examined by principal components analysis to explore the shape variations across three phases. The first two principal components (PCs) explain `r vari_two`% of variances, and the first three principal components (PCs) explain `r vari_three`% of variances. The PC1 relate to the height of the vessels from short to high, the PC2 reflects the neck and mouth constriction from wide to narrow, and the PC3 relates to the degree of flare of neck from curved to straight shape. The contribution of each principal component on the shape of vessels was listed in table (). The PC1 relative to the hight of the vessels accounts for nearly half of variances. Figure\@ref(fig:first-two-PCs-byplot) shows that a wider variation in pottery shape from the pre-European and post-European period compared to those from Chinese period.

```{r contribution-PCs}
# the contributions of each PC to the shapes
pc_contrib <- PCcontrib(op.p, nax = 1:3) 
pc_contrib$gg +  
  geom_polygon(fill="slategrey", col="black") + 
  ggtitle("Shape variation along PC axes") +
  theme_minimal(base_size = 6)

@@

# PC1 is height of vessel
# PC2 is neck constriction
# PC3 is height of neck and roundness of body
```

>>>>

`Post-European` <- mshp$poe %T>% coo_plot(border="blue")
`Pre-European` <-  mshp$pre %T>% coo_plot(border="red", plot.new=FALSE)
`Chinese` <-       mshp$chi %T>% coo_plot(border="green", plot.new=FALSE)
legend("topright", 
       lwd=1,
       col=c("red", "blue", "green"), 
       legend=c( "Pre-European", "Post-European", "Chinese"))
```

@@

# Doesn't work if the plot is not produced by ggplot?
# plot_grid(a, b, labels = c('A', 'B'))

@@

```{r PCA-table}
# these are fast
op.p <- PCA(op)
# proportion of variances explained, PC1 and PC2 explain 79% of variation
pro_vari <-scree_plot(op.p)$data
pc1 <- round(pro_vari$proportion[1],4)*100
pc3 <- round(pro_vari$proportion[3],4)*100
vari_two <- round(pro_vari$cumsum[2],4)*100
vari_three <- round(pro_vari$cumsum[3],4)*100
```

The elliptic Fourier coefficients are examined by principal components analysis to explore the shape variations across three phases. The first two principal components (PCs) explain `r vari_two`% of the total variances, of which `r pc1`% is explained by the first principle component. With the third component, the first three principal components explain `r vari_three`% of the total variances. Figure \@ref(fig:contribution-PCs) shows the shape variation associated to the first three principal components. From the negative to the positive scores, the PC1 represents the height of the vessels from tall to short and the roundness of the body from round to oval shaped. The PC2 relates to the neck and mouth constriction from narrow to wide. The PC3 explains a smaller portion of the variance (`r pc3`%), which relates to the degree of flare of neck from curved to straight shape. The first two components account for most of the variance in relation to three phases were represented in Figure \@ref(fig:first-two-PCs-byplot). The result reflects a large overlap of three groupings especially for shapes in the pre-European and post-European period that shows some extent of similarity and consistency. However, the spread of shape distribution indications a wider variation in shapes in the pre-European and post-European period compared to those in Chinese period along both PC1 and PC2 axes. In other words, there is a decreasing in shape variance in the Chinese period that shares the features of shorter height and narrower mouth.

```{r contribution-PCs, fig.cap= "Pottery shape variation based on scores along principal component axes. The rows from the top to the bottom represent the first to third principal component respectively. Each column shows the reconstructed shape variable according to the score mean ±2 standard deviation (SD) with a 0.5 interval"}
# the contributions of each PC to the shapes
pc_contrib <- PCcontrib(op.p, nax = 1:3) 
dev.off()
pc_contrib$gg +  
  geom_polygon(fill="slategrey", col="black") + 
  theme_minimal(base_size = 6)

@@

# PC1 is height of vessel
# PC2 is neck constriction
# PC3 is the degree of flare of neck
```
